Temporal ordering and registration of images in studies of developmental dynamics

### 2014-11-03 image_analysis_paper_development/manuscript.tex
added stuff to submit to development

%% PNAStmpl.tex
%% Template file to use for PNAS articles prepared in LaTeX
%% Version: Apr 14, 2008

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BASIC CLASS FILE
%% PNAStwo for two column articles is called by default.
%% Uncomment PNASone for single column articles. One column class
%% and style files are available upon request from pnas@nas.edu.
%% (uncomment means get rid of the '%' in front of the command)

%\documentclass{pnasone}
\documentclass{pnastwo}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Changing position of text on physical page:
%% Since not all printers position
%% the printed page in the same place on the physical page,
%% you can change the position yourself here, if you need to:

% \advance\voffset -.5in % Minus dimension will raise the printed page on the
                         %  physical page; positive dimension will lower it.

%% You may set the dimension to the size that you need.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL GRAPHICS STYLE FILE

%% Requires graphics style file (graphicx.sty), used for inserting
%% .eps files into LaTeX articles.
%% Note that inclusion of .eps files is for your reference only;
%% when submitting to PNAS please submit figures separately.

%% Type into the square brackets the name of the driver program
%% that you are using. If you don't know, try dvips, which is the
%% most common PC driver, or textures for the Mac. These are the options:

% [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
% [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
% [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]

%\usepackage[dvips]{graphicx}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL POSTSCRIPT FONT FILES

%% PostScript font files: You may need to edit the PNASoneF.sty
%% or PNAStwoF.sty file to make the font names match those on your system.
%% Alternatively, you can leave the font style file commands commented out
%% and typeset your article using the default Computer Modern
%% fonts (recommended). If accepted, your article will be typeset
%% at PNAS using PostScript fonts.

% Choose PNASoneF for one column; PNAStwoF for two column:
%\usepackage{PNASoneF}
\usepackage{PNAStwoF}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ADDITIONAL OPTIONAL STYLE FILES

%% The AMS math files are commonly used to gain access to useful features
%% like extended math fonts and math commands.

\usepackage{amssymb,amsfonts,amsmath}

%\usepackage{tikz}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL MACRO FILES
%% Insert self-defined macros here.
%% \newcommand definitions are recommended; \def definitions are supported

%\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
%\def\s{\sigma}

\newcommand{\SI}[0]{{\it SI Materials and Methods}}
\newcommand{\fig}[0]{Fig.}

\font\xfigtextfontit=\frutigeroblique at 7pt

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Don't type in anything in the following section:
%%%%%%%%%%%%
%% For PNAS Only:
\contributor{Submitted to Proceedings
of the National Academy of Sciences of the United States of America}
\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
\copyrightyear{2008}
\issuedate{Issue Date}
\volume{Volume}
\issuenumber{Issue Number}
%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% For titles, only capitalize the first letter
%% \title{Almost sharp fronts for the surface quasi-geostrophic equation}

%% Enter authors via the \author command.
%% Use \affil to define affiliations.
%% (Leave no spaces between author name and \affil command)

%% Note that the \thanks{} command has been disabled in favor of
%% a generic, reserved space for PNAS publication footnotes.

%% \author{<author name>
%% \affil{<number>}{<Institution>}} One number for each institution.
%% The same number should be used for authors that
%% are affiliated with the same institution, after the first time
%% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
%% Then, before last author ...
%% \and
%% \author{<author name>
%% \affil{<number>}{}}

%% For example, assuming Garcia and Sonnery are both affiliated with
%% Universidad de Murcia:
%% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
%% United Kingdom},
%% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
%% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}

\author{Carmeline~J.~Dsilva\affil{1}{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
Bomyi~Lim\affil{1}{},
Thomas~J.~Levario\affil{2}{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
Hang~Lu\affil{2}{},
Amit~Singer\affil{3}{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA},
Stanislav~Y.~Shvartsman\affil{1}{} \affil{5}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA},
\and
Ioannis~G.~Kevrekidis\affil{1}{} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}}

\contributor{Submitted to Proceedings of the National Academy of Sciences
of the United States of America}

\significancetext{
Spatial registration and temporal ordering of images of developing embryos are one of the first steps in the reconstruction of developmental dynamics. 
%
When the number of images is small and the differences between them are easily recognizable, image registration and ordering can be done manually; these tasks become challenging when the data sets are larger and more complex. 
%
Here we show this can be automated and combined in a single
step using recent advances in data mining. 
%
To illustrate this approach, we order and register data from imaging studies of {\xfigtextfontit  Drosophila} embryogenesis. }

%% The \maketitle command is necessary to build the title page.
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{article}

@@

Imaging studies provide unique insights into the dynamics of pattern formation and morphogenesis in developing tissues.
%
Developmental progress is commonly reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific developmental system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in  {\it Drosophila} embryos.

@@

%% When adding keywords, separate each term with a straight line: |
\keywords{temporal ordering | image registration | vector diffusion maps}

%\section{Significance}

%\vspace{0.5cm}

%% Optional for entering abbreviations, separate the abbreviation from
%% its definition with a comma, separate each pair with a semicolon:
%% for example:
%% \abbreviations{SAM, self-assembled monolayer; OTS,
%% octadecyltrichlorosilane}

% \abbreviations{}

%% The first letter of the article should be drop cap: \dropcap{}
%\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts

%% Enter the text of your article beginning here and ending before
%% \begin{acknowledgements}
%% Section head commands for your reference:
%% \section{}
%% \subsection{}
%% \subsubsection{}

\dropcap{I}n one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue. 
%
The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of the relevant process along its developmental trajectory \cite{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
%
Importantly, the ``age'' of any given embryo arrested in its development is often only approximately known; typically what is known is
a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
%
Furthermore, images are often collected in different spatial orientations.
%
In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered}, and then ordered in time.
%
%We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.

@@

%\begin{figure}[t]
%\includegraphics{fig1}
%\caption{Caricature illustrating the tasks of image registration and temporal ordering. {\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development. {\xfigtextfontit (B)} Registered and ordered samples. For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
%%\label{fig:fish}
%\customlabel{fig:fish}{1}
%\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
%\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
%\end{figure}

We present a robust algorithmic approach to registration and temporal ordering.
%
In contrast to a number of previous methodologies \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
%
Our approach is based on a manifold learning algorithm (vector diffusion maps \cite{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data} and classification of CT scans \cite{fernandez2014diffusion}.
%
Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \cite{jaeger2012drosophila}.
%
Our first data set, from a live imaging study where the correct rotational orientation and temporal order are independently known, will be used to validate our approach.
%
Our second data set consists of images from fixed embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.

\section{Results}

\subsection{Vector diffusion maps for registration and temporal ordering}

Vector diffusion maps \cite{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries (such as rotations of the images) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \cite{singer2011angular} for image registration and {\em diffusion maps} \cite{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing embryos with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
We assume that the main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying developmental dynamics.

Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{subfig:synch1}, where each image is represented as a vector, and the goal is to align the set of vectors.
%
We first compute the angles needed to align pairs of vectors (or images).  
%
In general, this requires no template function \cite{ahuja2007template} or image landmarks \cite{ian1998statistical}.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{subfig:synch2}.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can successfully register data sets even when many of the pairwise measurements are inaccurate \cite{singer2011angular}.

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%
Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{subfig:dmaps1}:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

%\begin{figure}[t]
%\includegraphics{fig2}
%\caption{Schematic illustrating angular synchronization and diffusion maps. {\xfigtextfontit (A)} Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. {\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}. {\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by darker (``stronger'') edges. {\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
%%\label{fig:schematics}
%\customlabel{fig:schematics}{2}
%\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
%\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
%\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
%\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
%\end{figure}

%\begin{figure*}[t]
%\includegraphics{fig3}
%\caption{Method validation using live imaging of {\xfigtextfontit Drosophila} embryos.{\xfigtextfontit (A)} Selected images from a live imaging study of a {\xfigtextfontit Drosophilla} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. {\xfigtextfontit (B)} Images from {\xfigtextfontit A} after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. {\xfigtextfontit (C)} The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle {\xfigtextfontit (D)} The correlation between the recovered rank (using vector diffusion maps) and the true rank. {\xfigtextfontit (E)} The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies. {\xfigtextfontit (F)} The median rank correlation coefficient (over $200$ samples) as a function of the number of images in the data set. }
%\customlabel{fig:movie}{3}
%\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}{\it A}}
%\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{\it B}}
%\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{\it C}}
%\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{\it D}}
%\customlabel{subfig:movie_error_table}{\ref{fig:movie}{\it E}}
%\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{\it F}}
%\end{figure*}

\subsection{Method validation using live imaging}

To validate the proposed approach, we applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
%
This data set was obtained through live imaging near the posterior pole of a vertically oriented {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
%
During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.

@@

In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \cite{singer2011angular}.
%
Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%
To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%
Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
%
\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%
As expected, the accuracy in the recovered ordering increases with the number of images. 
% 
At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%
Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
%
In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

\subsection{Data sets with interembryo variability}

@@

Finally, we applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
%
\fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
%
This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provides information about tissue morphology and the spatial distribution of two regulatory proteins.
%
Each image shows an optical cross-section of a {\em different} vertically oriented embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
%
The nuclei (gray) were labeled with DAPI, a DNA stain.
%
Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \cite{Lim2013kinetics}.

>>>>

\documentclass[twocolumn, 10pt]{article}

\usepackage[margin=0.75in]{geometry}
\usepackage{amssymb,amsfonts,amsmath}
\usepackage{graphicx}

\usepackage{natbib}

\renewcommand{\figurename}{Fig.}
\newcommand{\SI}[0]{\textit{SI Materials and Methods}}
\newcommand{\fig}[0]{Fig.}

\hyphenpenalty=10000

@@

\author{Carmeline~J.~Dsilva \thanks{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
Bomyi~Lim \footnotemark[1],
Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA},
Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}, \\
and Ioannis~G.~Kevrekidis \footnotemark[1] \footnotemark[4]}

\date{}

@@

\maketitle

@@

Imaging studies provide unique insights into the dynamics of pattern formation and morphogenesis in developing tissues.
%
Developmental progress is commonly reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific developmental system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in  \textit{Drosophila} embryos.

\vspace{1em}

\noindent
KEY WORDS: temporal ordering, image registration, vector diffusion maps

@@

\section*{Introduction}

In one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue. 
%
The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of the relevant process along its developmental trajectory \citep{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
%
Importantly, the ``age'' of any given embryo arrested in its development is often only approximately known; typically what is known is
a certain time window to which a collection of embryos belongs \citep{ng2012large, richardson2014emage, castro2009automatic}.
%
Furthermore, images are often collected in different spatial orientations.
%
In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered}, and then ordered in time.
%
%We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.

@@

\begin{figure}[t]
\includegraphics{fig1}
\caption{Caricature illustrating the tasks of image registration and temporal ordering. (A) Images of ``samples'', each in a different orientation and a different stage of development. (B) Registered and ordered samples. For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
\end{figure}

We present a robust algorithmic approach to registration and temporal ordering.
%
In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
%
Our approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
%
Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our first data set, from a live imaging study where the correct rotational orientation and temporal order are independently known, will be used to validate our approach.
%
Our second data set consists of images from fixed embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.

\section*{Results}

\subsection*{Vector diffusion maps for registration and temporal ordering}

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries (such as rotations of the images) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing embryos with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
We assume that the main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying developmental dynamics.

\begin{figure}[t]
\includegraphics{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. (A) Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. (B) The vectors from A, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in A: the differences between a pair of angles in B is the same as the pairwise angle in A. (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \SI), so that close data points are connected by darker (``stronger'') edges. (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
%\label{fig:schematics}
\customlabel{fig:schematics}{2}
\customlabel{subfig:synch1}{\ref{fig:schematics}A}
\customlabel{subfig:synch2}{\ref{fig:schematics}B}
\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
\end{figure}

Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{subfig:synch1}, where each image is represented as a vector, and the goal is to align the set of vectors.
%
We first compute the angles needed to align pairs of vectors (or images).  
%
In general, this requires no template function \citep{ahuja2007template} or image landmarks \citep{ian1998statistical}.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{subfig:synch2}.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can successfully register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%
Such a manifold can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{subfig:dmaps1}:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

\begin{figure*}[t]
\includegraphics{fig3}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies. (F) The median rank correlation coefficient (over $200$ samples) as a function of the number of images in the data set. }
\customlabel{fig:movie}{3}
\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}A}
\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{B}}
\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{C}}
\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{D}}
\customlabel{subfig:movie_error_table}{\ref{fig:movie}{E}}
\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{F}}
\end{figure*}

\subsection*{Method validation using live imaging}

To validate the proposed approach, we applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
%
This data set was obtained through live imaging near the posterior pole of a vertically oriented {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
%
During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \citep{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.

@@

In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
%
Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%
To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%
Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
%
\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%
As expected, the accuracy in the recovered ordering increases with the number of images. 
% 
At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%
Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
%
In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

\subsection*{Data sets with interembryo variability}

@@

Finally, we applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
%
\fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
%
This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provides information about tissue morphology and the spatial distribution of two regulatory proteins.
%
Each image shows an optical cross-section of a {\em different} vertically oriented embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
%
The nuclei (gray) were labeled with DAPI, a DNA stain.
%
Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \citep{Lim2013kinetics}.

\begin{figure*}[t]
\includegraphics{fig4}
\caption{Analyzing images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.97$. }
\customlabel{fig:data2}{4}
\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{A}}
\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{B}}
\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{C}}
\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{D}}
\end{figure*}
